\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-04-10 forwards-and-backwards.tex
minor fixen

\comment{TODO: cite \citet{lam2023introduction,cantrell2004spatial}}

@@

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we expect that if we take a second limit of $\epsilon \to 0$,
thus scaling the kernels appearing in $r$, $\gamma$, and $F$ and making interactions pointwise,
we should recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which we 
retain a signal of our two stage approach to modelling in the nonlinear diffusion term
and the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.

@@

We shall write $\mu_\theta(x, X/N)$ for the per-capita death rate of mature individuals
in the population.
In order for the 
population density to change over timescales of order $\theta$, we should like the net 
per capita reproductive rate 
to scale as $1/\theta$. 
In classical models, in which $r$, $\gamma$, and $\mu$ are constant, 
this quantity is simply $r\gamma -\mu$. 
Here, because production of juveniles and their establishment are
mediated by population density measured relative to different points, 
the net reproductive rate will take a more complicated form. 
In particular, the total rate of production of mature offspring by an individual at $x$ will be
\[
	\gamma\big(x,\smooth{\gamma}X(x)/N\big)\int r\big(y, \smooth{r}X(y)/N\big)q_{\theta}(x,dy).
\]
Nonetheless, it will be convenient to define the death rate $\mu_\theta$ in terms of 
its deviation from $r\gamma$. 
To this end, we define
the death rate of an individual at $x$ to be
\begin{align} \label{eqn:mu_defn}
    \mu_\theta(x)
    =
    \max\left\{0, 
        r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N)
    \right\} ,
\end{align}
where this equation defines $F(x, m) : \IR^d \times \IR_{\ge 0} \to \IR$,
and $\rho_F$ is again a smooth kernel. 
\comment{FIX UP THIS, REFER FORWARD}
The function $F$ is nearly
the net per capita reproductive rate, scaled by $\theta$,
and would be equal to it in a nonspatial model;
but 
It will be convenient to work with $F$ instead of $\mu$.

@@

\begin{corollary} \label{cor:superprocess_uniqueness}
    Under the assumptions of Theorem~\ref{thm:nonlocal_convergence},
    if the martingale problem
    defined by equations~\eqref{eqn:limiting_mgale_problem} 
and~\eqref{eqn:limiting_mgale_variation}
    has a unique solution,
    then $(\eta^N_t)_{t \ge 0}$ converges weakly
to that solution
    as $N \to \infty$.
\end{corollary}
When $\alpha>0$, the limit points can be thought of as interacting superprocesses. For
example, when $r$ and $\gamma$ are constant, and 
$F(x,\eta_s)=1-\smooth{F} \eta_s(x)$,
we recover a superprocess with nonlinear death rates corresponding to logistic growth
\comment{cite Etheridge 2004}
that is a continuous limit of the Bolker-Pacala model \citep{bolker/pacala:1997,bolker/pacala:1999}.
We are not aware of a general result to determine when we will have uniqueness of 
solutions to the martingale problem of Theorem~\ref{thm:nonlocal_convergence}
when $\alpha>0$.
However, the Dawson--Girsanov transform tells us that we have uniqueness in this special 
case of the superprocess with nonlinear death rates, and 
Perkins stochastic calculus (and its adaptation to a 
lookdown setting) provides uniqueness for cases with interactions in the 
dispersal mechanism of
the superprocess. We refer 
to~\cite{dawson:1993, perkins:1992}, and \cite{donnelly/kurtz:1999} 
for approaches to showing that the martingale problem is well-posed in the case $\alpha>0$.

@@

In one dimension, equation~(\ref{PME}) has an explicit travelling wave solution
\begin{align} \label{eqn:pme_wave}
    w^P(t, x)
    :=
    \left( 1 - e^{ \frac{1}{2} (x - x_0 - t) } \right)_+ .
\end{align}
Notice that the wave profile has a sharp boundary at $x = x_0 + t$.
There are also travelling wave solutions with $c>1$ \citep{gilding/kersner:2005},
which lack this property.
However, for initial conditions that decay sufficiently rapidly at infinity,
such as one might use in modelling a population invading new territory,
the solution converges to \eqref{eqn:pme_wave} \citep{kamin/rosenau:2004}.
In Figure~\ref{fig:pme_waves} we show simulations of the individual based model
described above. They
display travelling wave solutions
similar to numerical solutions of~(\ref{PME}),
with increasingly good agreement for larger $\theta$ and $N$.
As expected, noise slows down the wave.
\comment{Not sure that noise will slow things drastically because the 
biggest production of
mass happens behind the wavefront.}

@@

Can these two situations be distinguished from summaries of lineage movement?
The first has lineage generator
\[
    f \mapsto \Lgen f = r \gamma \left( \Delta f + 2 \grad \log(\gamma w) \cdot \grad f \right),
\]
while the second has lineage generator $f \mapsto \lambda(x) \Lgen f(x)$.
In other words,
although the stationary profile of the population is unchanged when we scale
local establishment and death by $\lambda$,
the motion of lineages is sped up locally by $\lambda$.
This corresponds to making areas with $\lambda > 1$ more ``sink-like'' 
and $\lambda < 1$ more ``source-like'':
if $\lambda(x) > 1$, then at $x$ both the death rate and probability of 
establishment of new individuals are higher.
As a result, lineages in the second model spend more time in areas with $\lambda < 1$,
i.e., those areas have higher reproductive value, %long-term fitness
something that is, in principle, discernible from genetic data.
\comment{clarify why: smaller $N_e$}

@@

Write $\tau_a$ for the birth time of the label $a$ and 
$\sigma_a = \lim_{K \to \infty} \inf\{t \ge 0: U_a(t) > K\}$
the time the level hits $\infty$.
Suppose that the spatial locations and level of each line of descent $a$
solve, for $\tau_a \le t < \sigma_a$,
\begin{align*}
X_a(t)
    &=
    X_a(\tau_a)
    + \int_{\tau_a}^{t}
        \beta(X_a(s), \eta_s) ds
    + \int_{\tau_a}^{t}
        J(X_a(s),\eta_s) dW_a(s)
    \\
U_a(t)
    &=
    U_a(\tau_a)
    + \int_{\tau_a}^{t}
    \bigg(
        \alpha \gamma(X_a(s),\eta_s)
        r(X_a(s), \eta_s) U_a(s)^2
\\ &\qquad \qquad \qquad {}   
        -
        \big\{
            \gamma(X_a(s),\eta_s) \DG r(X_a(s),\eta_s)
            + F(X_a(s), \eta_s)
        \big\}
        U_a(s)
    \bigg)
    ds .
\end{align*}
Each point in each $R_a$ denotes a potential birth time for $a$:
if the $j^\text{th}$ point in $R_a$ is $(\tau, v)$, with 
$\tau_a \le \tau < \sigma_a$,
then a new line of descent with label $a \concat j$ is produced,
with birth time $\tau_{a \concat j} = \tau$,
    location $X_{a \concat j}(\tau) = X_a(\tau)$, and level
$$
    U_{a \concat j}(\tau) = U_a(\tau)
    + \frac{v}{ 2 \alpha \gamma(X_a(\tau), \eta_\tau) r(X_a(\tau), \eta_\tau) } ,
$$
if this is finite.
For any solution to the equations above, the processes defined by
\begin{align*}
    \eta_t
    =
    \lim_{u_0 \to \infty} \frac{1}{u_0}
        \sum_{a : \tau_a \le t < \sigma_a \; ; \; U_a(t) < u_0} \delta_{X_a(t)}
    \qquad \text{and} \qquad
    \lp_t = \sum_{a : \tau_a \le t < \sigma_a} \delta_{(X_a(t), U_a(t))} 
\end{align*}
are solutions to the martingale problems of
Theorems~\ref{thm:nonlocal_convergence} and~\ref{thm:lookdown_convergence},
respectively.
\end{proposition}

@@

\begin{proof}[of Proposition~\ref{prop:limiting_construction}]
    The fact that a solution to the system of equations
    \comment{label the system and refer to it}
    is a solution to the martingale problem of Theorem~\ref{thm:lookdown_convergence}
    is an application of It\^o's theorem.
    Furthermore, in Proposition~\ref{thm:mmt_application} we showed that
    the conditional Poisson property of $\lp_0$ is preserved
    (i.e., holds for $\lp_t$ for all $t$), and so
    $(\eta_t)_{t \ge 0}$ is well-defined,
    and furthermore that $\eta_t$ is a solution
    to the martingale problem of Theorem~\ref{thm:nonlocal_convergence}.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proofs}
\label{sec:proofs}

\comment{this section (6) is no long just "proofs" since they're in section 7 and 8 too}
\comment{make sure this reads okay and describes just this section (6); the end-of-the-intro paragraph describes structure}
In the following sections we present formal proofs of our three scaling limits.
In Subsection~\ref{sec:population_density_proof} 
we prove Theorem~\ref{thm:nonlocal_convergence}, to obtain a limit in
which interactions between individuals in the population are nonlocal.
In Subsection~\ref{subsec:nonlocal to local}
we show how, in two important examples in which the nonlocal
limit is respectively a deterministic solution to a non-local equation of 
reaction diffusion type and a deterministic solution to a nonlocal 
porous medium equation with an additional logistic growth term, 
one can pass to a further 
limit to obtain a classical PDE.
In Section~\ref{subsec:one step convergence proof}
we prove our third scaling limit, in which 
we pass from our population model 
directly to a reaction diffusion equation without going via a nonlocal limit.
Finally, in Section~\ref{sec:lookdown_proofs} we prove our results concerning both convergence of the lookdown representation
and characterization of the lineage process in the deterministic case.

@@

Below we will have frequent use for the quantity
\comment{perhaps $B$ should have a superscript $\theta$}
\begin{equation}
\label{generator Bf}
    B_f(x, \eta) = \theta \int_{\IR^d} (f(y) r(y, \eta) - f(x) r(x, \eta)) q_\theta(x, dy) .
\end{equation}

@@

\begin{proof}[of Lemma~\ref{lem:conditions on r}]
Here, we need to prove that $|\gamma(x, \eta) B_1(x, \eta)|$ is bounded,
uniformly over $x$ and $\eta$.
First suppose that
Condition~\ref{control through r} of  
Lemma~\ref{lem:conditions on r} is satisfied.
We write
\begin{align*}
    r\big(y,\rho_r*\eta(y)\big) - r\big(x,\rho_r*\eta(x)\big)
    &=
    \sum_i (y_i - x_i) \partial_{x_i} \Big(r\big(x,\rho_r*\eta(x)\big)\Big) 
    \\ &\qquad {}
    +
    \frac{1}{2} \sum_{ij} (y_i - x_i)(y_j - x_j)
        \partial_{x_i x_j} \Big(r\big(\smooth{r}\eta(z)\big)\Big) ,
\end{align*}
where $z = \kappa x + (1-\kappa) y$ for some $0 \le \kappa \le 1$.
Condition~\ref{control through r} 
gives uniform bounds on the derivatives of $r(x, \rho_r*\eta(x))$ 
in this expression and so we have a bound of the form
\begin{multline*}
\Big| r\big(y,\rho_r*\eta(y)\big) - r\big(x,\rho_r*\eta(x)\big) \Big|
\\
\leq
        K_1 \left|
            \theta \int_{\IR^d} \sum_i (y_i - x_i) q_\theta(x, dy)
        \right|
        +
        K_2 \left|
            \theta \int_{\IR^d} \sum_{ij} (y_i - x_i) (y_j - x_j)  q_\theta(x, dy)
        \right| 
\end{multline*}
for suitable constants $K_1$, $K_2$.
This is bounded uniformly because
$q_\theta(x, dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
and covariance $\covq(x)/\theta$, and both $\meanq(x)$ and $\covq(x)$
are bounded uniformly in $x$ under Assumptions~\ref{def:model_setup}.
\comment{typo}
u implies that $|B_1|$ is uniformly bounded;
since $\gamma$ is bounded as well, so is $|\gamma B_1|$.

Now suppose that instead
Condition~\ref{control through gamma} of Lemma~\ref{lem:conditions on r} is satisfied.
is satisfied.
First note that
\begin{multline*}
    r(x, \smooth{r}\eta(y)) - r(x, \smooth{r}\eta(x))
    \\
=
    (\smooth{r}\eta(y) - \smooth{r}\eta(x)) r'(x,\smooth{r}\eta(x))
    + \frac{1}{2} (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2 r''(x, \overline{m}),
\end{multline*}
where $\overline{m}= \kappa'\smooth{r}\eta(x) +(1-\kappa') 
\smooth{r}\eta(y)$ for some $0\leq \kappa'\leq 1$, and
we have used $r'$ and $r''$ to denote the first and second derivatives of $r(x,m)$
with respect to the second argument.
So, in this case
\begin{align*}
&
\left| \theta 
    \int_{\IR^n} 
        \big\{
            r\big(x,\smooth{r}\eta(y)\big)
            -
            r\big(x,\smooth{r}\eta(x)\big)
        \big\}
    q_{\theta}(x,dy) 
\right| 
\\ &\qquad
\le
   \|r'\|_\infty
    \left|
        \theta \int_{\IR^d}
            (\smooth{r}\eta(y) - \smooth{r}\eta(x))
        q_\theta(x, dy)
    \right|
    +
   \|r''\|_\infty
    \left|
        \theta \int_{\IR^d}
            (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2
        q_\theta(x, dy)
    \right| .
\end{align*}
Therefore,
under Condition~\ref{control through gamma}, $|B_1(x,\eta)|$ is bounded by 
a constant times $\smooth{\gamma}\eta(x)+(\smooth{\gamma}\eta(x))^2$;
since $m^2\gamma(m)$ is bounded, so is $|\gamma(x,\eta) B_1(x,\eta)|$.
\end{proof}

@@

\begin{proof}
Consider the semimartingale decomposition from equation~\eqref{eqn:eta_martingale}:
\begin{align} \label{eqn:eta_f_mgale_decomp}
    \langle f, \eta^N_t \rangle
    &=
    \langle f, \eta^N_0 \rangle
    + \int_0^t \int_{\IR^d} \big\{
        \gamma(x, \eta^N_s) B_f(x, \eta^N_s)
        + f(x) F(x, \eta^N_s)
        \big\} \eta^N_s(dx) ds
    + M^N_t(f) ,
\end{align}
where $M^N_t(f)$ is a martingale and $B_f$ is defined in \ref{generator Bf}.

>>>>

@@

By replacing $\rho$ by
$\rho^\epsilon(\cdot)=\rho(\cdot/\epsilon)/\epsilon^d$, 
we can also scale the ``width'' of the 
region over which we measure local population density. 
When the population follows~(\ref{general deterministic limit}), 
we expect that if we take a second limit of $\epsilon \to 0$,
thus scaling the kernels appearing in $r$, $\gamma$, and $F$ and making interactions pointwise,
we should recover a nonlinear PDE.
We verify that this is indeed the case in two important examples:
a special case of the porous medium equation with a logistic growth term, in which we 
retain a signal of our two stage approach to modelling in the nonlinear diffusion term
and the limiting equation takes the form 
\begin{equation}
	\label{PME1}
	\partial_t \varphi = \Delta (\varphi^2)+\varphi(1-\varphi);
\end{equation}
and a wide class of semi-linear PDEs of the form 
\begin{equation}
\label{semilinear PDE}
\partial_t \varphi = \DG^*\varphi+ \varphi F(\varphi),
\end{equation}
which includes the Fisher-KPP equation and the Allen-Cahn equation.
Equations of this form have been studied extensively
in the context of spatial ecology;
see for instance \citet{lam2023introduction} and \citet{cantrell2004spatial}.

@@

We shall write $\mu_\theta(x, X/N)$ for the per-capita death rate of mature individuals
in the population.
In order for the 
population density to change over timescales of order $\theta$, we should like the net 
per capita reproductive rate 
to scale as $1/\theta$. 
In classical models, in which $r$, $\gamma$, and $\mu$ are constant, 
this quantity is simply $r\gamma -\mu$. 
Here, because production of juveniles and their establishment are
mediated by population density measured relative to different points, 
the net reproductive rate will take a more complicated form. 
In particular, the total rate of production of mature offspring by an individual at $x$ will be
\[
	\gamma\big(x,\smooth{\gamma}X(x)/N\big)\int r\big(y, \smooth{r}X(y)/N\big)q_{\theta}(x,dy).
\]
Nonetheless, it will be convenient to define the death rate $\mu_\theta$ in terms of 
its deviation from $r\gamma$. 
To this end, we define
the death rate of an individual at $x$ to be
\begin{align} \label{eqn:mu_defn}
    \mu_\theta(x)
    =
    \max\left\{0, 
        r(x, \smooth{r} X(x) / N) \gamma(x, \smooth{\gamma} X(x) / N)
        - \frac{1}{\theta} F(x, \smooth{F} X(x) / N)
    \right\} ,
\end{align}
where this equation defines $F(x, m) : \IR^d \times \IR_{\ge 0} \to \IR$,
and $\rho_F$ is again a smooth kernel. 
The function $F$ is nearly
the net per capita reproductive rate, scaled by $\theta$,
and would be equal to it in a nonspatial model;
but, as discussed later,
differs because offsprings' establishment probability is measured at their new location
rather than their parents', and so differs from $r(x)$.
For the most part, we work with $F$ instead of $\mu$.

@@

\begin{corollary} \label{cor:superprocess_uniqueness}
    Under the assumptions of Theorem~\ref{thm:nonlocal_convergence},
    if the martingale problem
    defined by equations~\eqref{eqn:limiting_mgale_problem} 
and~\eqref{eqn:limiting_mgale_variation}
    has a unique solution,
    then $(\eta^N_t)_{t \ge 0}$ converges weakly
to that solution
    as $N \to \infty$.
\end{corollary}
When $\alpha>0$, the limit points can be thought of as interacting superprocesses. For
example, when $r$ and $\gamma$ are constant, and 
$F(x,\eta_s)=1-\smooth{F} \eta_s(x)$,
we recover a superprocess with nonlinear death rates corresponding to logistic growth
\citep{etheridge:2004}
that is a continuous limit of the Bolker-Pacala model \citep{bolker/pacala:1997,bolker/pacala:1999}.
We are not aware of a general result to determine when we will have uniqueness of 
solutions to the martingale problem of Theorem~\ref{thm:nonlocal_convergence}
when $\alpha>0$.
However, the Dawson--Girsanov transform tells us that we have uniqueness in this special 
case of the superprocess with nonlinear death rates, and 
Perkins stochastic calculus (and its adaptation to a 
lookdown setting) provides uniqueness for cases with interactions in the 
dispersal mechanism of
the superprocess. We refer 
to~\cite{dawson:1993, perkins:1992}, and \cite{donnelly/kurtz:1999} 
for approaches to showing that the martingale problem is well-posed in the case $\alpha>0$.

@@

In one dimension, equation~(\ref{PME}) has an explicit travelling wave solution
\begin{align} \label{eqn:pme_wave}
    w^P(t, x)
    :=
    \left( 1 - e^{ \frac{1}{2} (x - x_0 - t) } \right)_+ .
\end{align}
Notice that the wave profile has a sharp boundary at $x = x_0 + t$.
There are also travelling wave solutions with $c>1$ \citep{gilding/kersner:2005},
which lack this property.
However, for initial conditions that decay sufficiently rapidly at infinity,
such as one might use in modelling a population invading new territory,
the solution converges to \eqref{eqn:pme_wave} \citep{kamin/rosenau:2004}.
In Figure~\ref{fig:pme_waves} we show simulations of the individual based model
described above. They
display travelling wave solutions
similar to numerical solutions of~(\ref{PME}),
with increasingly good agreement for larger $\theta$ and $N$.
\comment{TODO: does it?}
As expected, noise slows down the wave.
\comment{Not sure that noise will slow things drastically because the 
biggest production of
mass happens behind the wavefront.}

@@

Can these two situations be distinguished from summaries of lineage movement?
The first has lineage generator
\[
    f \mapsto \Lgen f = r \gamma \left( \Delta f + 2 \grad \log(\gamma w) \cdot \grad f \right),
\]
while the second has lineage generator $f \mapsto \lambda(x) \Lgen f(x)$.
In other words,
although the stationary profile of the population is unchanged when we scale
local establishment and death by $\lambda$,
the motion of lineages is sped up locally by $\lambda$.
This corresponds to making areas with $\lambda > 1$ more ``sink-like'' 
and $\lambda < 1$ more ``source-like'':
if $\lambda(x) > 1$, then at $x$ both the death rate and probability of 
establishment of new individuals are higher.
As a result, lineages in the second model spend more time in areas with $\lambda < 1$,
i.e., those areas have higher reproductive value, %long-term fitness
something that is, in principle, discernible from genetic data
(because, for instance, making reproductive value less evenly distributed
reduces long-term genetic diversity).

@@

Write $\tau_a$ for the birth time of the label $a$ and 
$\sigma_a = \lim_{K \to \infty} \inf\{t \ge 0: U_a(t) > K\}$
the time the level hits $\infty$.
Suppose that the spatial locations and level of each line of descent $a$
solve, for $\tau_a \le t < \sigma_a$,
\begin{align}
    \label{eqn:limiting_construction}
    \begin{split}
X_a(t)
    &=
    X_a(\tau_a)
    + \int_{\tau_a}^{t}
        \beta(X_a(s), \eta_s) ds
    + \int_{\tau_a}^{t}
        J(X_a(s),\eta_s) dW_a(s)
    \\
U_a(t)
    &=
    U_a(\tau_a)
    + \int_{\tau_a}^{t}
    \bigg(
        \alpha \gamma(X_a(s),\eta_s)
        r(X_a(s), \eta_s) U_a(s)^2
\\ &\qquad \qquad \qquad {}   
        -
        \big\{
            \gamma(X_a(s),\eta_s) \DG r(X_a(s),\eta_s)
            + F(X_a(s), \eta_s)
        \big\}
        U_a(s)
    \bigg)
    ds .
    \end{split}
\end{align}
Each point in each $R_a$ denotes a potential birth time for $a$:
if the $j^\text{th}$ point in $R_a$ is $(\tau, v)$, with 
$\tau_a \le \tau < \sigma_a$,
then a new line of descent with label $a \concat j$ is produced,
with birth time $\tau_{a \concat j} = \tau$,
    location $X_{a \concat j}(\tau) = X_a(\tau)$, and level
$$
    U_{a \concat j}(\tau) = U_a(\tau)
    + \frac{v}{ 2 \alpha \gamma(X_a(\tau), \eta_\tau) r(X_a(\tau), \eta_\tau) } ,
$$
if this is finite.
For any solution to the equations above, the processes defined by
\begin{align*}
    \eta_t
    =
    \lim_{u_0 \to \infty} \frac{1}{u_0}
        \sum_{a : \tau_a \le t < \sigma_a \; ; \; U_a(t) < u_0} \delta_{X_a(t)}
    \qquad \text{and} \qquad
    \lp_t = \sum_{a : \tau_a \le t < \sigma_a} \delta_{(X_a(t), U_a(t))} 
\end{align*}
are solutions to the martingale problems of
Theorems~\ref{thm:nonlocal_convergence} and~\ref{thm:lookdown_convergence},
respectively.
\end{proposition}

@@

\begin{proof}[of Proposition~\ref{prop:limiting_construction}]
    The fact that a solution to the system of equations~\eqref{eqn:limiting_construction}
    is a solution to the martingale problem of Theorem~\ref{thm:lookdown_convergence}
    is an application of It\^o's theorem.
    Furthermore, in Proposition~\ref{thm:mmt_application} we showed that
    the conditional Poisson property of $\lp_0$ is preserved
    (i.e., holds for $\lp_t$ for all $t$), and so
    $(\eta_t)_{t \ge 0}$ is well-defined,
    and furthermore that $\eta_t$ is a solution
    to the martingale problem of Theorem~\ref{thm:nonlocal_convergence}.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proofs of convergence for nonlocal models}
\label{sec:proofs}

In the following sections we present formal proofs of our three scaling limits.
In Subsection~\ref{sec:population_density_proof} 
we prove Theorem~\ref{thm:nonlocal_convergence}, to obtain a limit in
which interactions between individuals in the population are nonlocal.
In Subsection~\ref{subsec:nonlocal to local}
we show how, in two important examples in which the nonlocal
limit is respectively a deterministic solution to a non-local equation of 
reaction diffusion type and a deterministic solution to a nonlocal 
porous medium equation with an additional logistic growth term, 
one can pass to a further 
limit to obtain a classical PDE.
In subsequent sections (Sections~\ref{subsec:one step convergence proof} and~\ref{sec:lookdown_proofs})
we prove our third scaling limit, in which 
we pass from our population model 
directly to a reaction diffusion equation without going via a nonlocal limit,
and finally our results concerning both convergence of the lookdown representation
and characterization of the lineage process in the deterministic case.
\comment{TODO: revisit this paragraph}

@@

Below we will have frequent use for the quantity
\begin{equation}
\label{generator Bf}
    B^\theta_f(x, \eta) = \theta \int_{\IR^d} (f(y) r(y, \eta) - f(x) r(x, \eta)) q_\theta(x, dy) .
\end{equation}

@@

\begin{proof}[of Lemma~\ref{lem:conditions on r}]
Here, we need to prove that $|\gamma(x, \eta) B^\theta_1(x, \eta)|$ is bounded,
uniformly over $x$ and $\eta$.
First suppose that
Condition~\ref{control through r} of  
Lemma~\ref{lem:conditions on r} is satisfied.
We write
\begin{align*}
    r\big(y,\rho_r*\eta(y)\big) - r\big(x,\rho_r*\eta(x)\big)
    &=
    \sum_i (y_i - x_i) \partial_{x_i} \Big(r\big(x,\rho_r*\eta(x)\big)\Big) 
    \\ &\qquad {}
    +
    \frac{1}{2} \sum_{ij} (y_i - x_i)(y_j - x_j)
        \partial_{x_i x_j} \Big(r\big(\smooth{r}\eta(z)\big)\Big) ,
\end{align*}
where $z = \kappa x + (1-\kappa) y$ for some $0 \le \kappa \le 1$.
Condition~\ref{control through r} 
gives uniform bounds on the derivatives of $r(x, \rho_r*\eta(x))$ 
in this expression and so we have a bound of the form
\begin{multline*}
\Big| r\big(y,\rho_r*\eta(y)\big) - r\big(x,\rho_r*\eta(x)\big) \Big|
\\
\leq
        K_1 \left|
            \theta \int_{\IR^d} \sum_i (y_i - x_i) q_\theta(x, dy)
        \right|
        +
        K_2 \left|
            \theta \int_{\IR^d} \sum_{ij} (y_i - x_i) (y_j - x_j)  q_\theta(x, dy)
        \right| 
\end{multline*}
for suitable constants $K_1$, $K_2$.
This is bounded uniformly because
$q_\theta(x, dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
and covariance $\covq(x)/\theta$, and both $\meanq(x)$ and $\covq(x)$
are bounded uniformly in $x$ under Assumptions~\ref{def:model_setup}.
\comment{typo}
u implies that $|B^\theta_1|$ is uniformly bounded;
since $\gamma$ is bounded as well, so is $|\gamma B^\theta_1|$.

Now suppose that instead
Condition~\ref{control through gamma} of Lemma~\ref{lem:conditions on r} is satisfied.
is satisfied.
First note that
\begin{multline*}
    r(x, \smooth{r}\eta(y)) - r(x, \smooth{r}\eta(x))
    \\
=
    (\smooth{r}\eta(y) - \smooth{r}\eta(x)) r'(x,\smooth{r}\eta(x))
    + \frac{1}{2} (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2 r''(x, \overline{m}),
\end{multline*}
where $\overline{m}= \kappa'\smooth{r}\eta(x) +(1-\kappa') 
\smooth{r}\eta(y)$ for some $0\leq \kappa'\leq 1$, and
we have used $r'$ and $r''$ to denote the first and second derivatives of $r(x,m)$
with respect to the second argument.
So, in this case
\begin{align*}
&
\left| \theta 
    \int_{\IR^n} 
        \big\{
            r\big(x,\smooth{r}\eta(y)\big)
            -
            r\big(x,\smooth{r}\eta(x)\big)
        \big\}
    q_{\theta}(x,dy) 
\right| 
\\ &\qquad
\le
   \|r'\|_\infty
    \left|
        \theta \int_{\IR^d}
            (\smooth{r}\eta(y) - \smooth{r}\eta(x))
        q_\theta(x, dy)
    \right|
    +
   \|r''\|_\infty
    \left|
        \theta \int_{\IR^d}
            (\smooth{r}\eta(y) - \smooth{r}\eta(x))^2
        q_\theta(x, dy)
    \right| .
\end{align*}
Therefore,
under Condition~\ref{control through gamma}, $|B^\theta_1(x,\eta)|$ is bounded by 
a constant times $\smooth{\gamma}\eta(x)+(\smooth{\gamma}\eta(x))^2$;
since $m^2\gamma(m)$ is bounded, so is $|\gamma(x,\eta) B^\theta_1(x,\eta)|$.
\end{proof}

@@

\begin{proof}
Consider the semimartingale decomposition from equation~\eqref{eqn:eta_martingale}:
\begin{align} \label{eqn:eta_f_mgale_decomp}
    \langle f, \eta^N_t \rangle
    &=
    \langle f, \eta^N_0 \rangle
    + \int_0^t \int_{\IR^d} \big\{
        \gamma(x, \eta^N_s) B^\theta_f(x, \eta^N_s)
        + f(x) F(x, \eta^N_s)
        \big\} \eta^N_s(dx) ds
    + M^N_t(f) ,
\end{align}
where $M^N_t(f)$ is a martingale and $B^\theta_f$ is defined in \ref{generator Bf}.

We need a bound on $\gamma(x,\eta)B^\theta_f(x,\eta)+fF$.
First note that we may rewrite $B^\theta_f(x, \eta)$ as
\begin{align}
    \nonumber
&
    \left| \theta 
        \int_{\IR^n} 
            \big\{
                f(y) r\big(y,\smooth{r}\eta(y)\big)
                -
                f(x) r\big(x, \smooth{r}\eta(x)\big)
            \big\}
        q_{\theta}(x,dy)
    \right|
\\ 
    &\qquad {} =
    \Big| \theta 
        \int_{\IR^n} 
            ( f(y) - f(x) )
            r\big(y, \smooth{r}\eta(y)\big)
        q_{\theta}(x,dy)
\\ 
\nonumber
    &\qquad \qquad {}  +
    \theta 
            f(x)
        \int_{\IR^n} 
            \big\{
                r\big(y,\smooth{r}\eta(y)\big)
                -
                r\big(x,\smooth{r}\eta(x)\big)
            \big\}
        q_{\theta}(x,dy) 
    \Big| 
\\ 
    &\qquad {} \leq
    \left| \theta 
        \int_{\IR^n} 
            ( f(y) - f(x) )
            r(y, \smooth{r}\eta(y))
        q_{\theta}(x,dy)
    \right|
    +
    \|f\|_\infty |B^\theta_1(x, \eta)| .
\end{align}
\comment{TODO FIX THIS}
Using the intermediate value theorem,
then the fact that $0 \le r \le 1$,
there is a $\kappa$ in $[0, 1]$ such that the first term is
\begin{align*}
    &
    \left| \theta 
        \int_{\IR^n} 
            ( f(y) - f(x) )
            r(y, \eta)
        q_{\theta}(x,dy)
    \right|
    \\ &\qquad =
    \left| \theta 
        \int_{\IR^n} 
            \left(
                \sum_i (y-x)_i \partial_{x_i} f(x)
                +
                \frac{1}{2} \sum_{ij} (y-x)_i (y-x)_j \partial_{x_i x_j} f(x + \kappa(y-x))
            \right)
            r(y, \eta)
        q_{\theta}(x,dy)
    \right|
    \\ &\qquad \leq
    \|f'\|_\infty \left| \theta 
        \int_{\IR^n} 
            (y-x) r(y, \eta)
        q_{\theta}(x,dy)
    \right|
    + \frac{1}{2} \|f''\|_\infty \left| \theta 
        \int_{\IR^n} 
            (y-x)^2
        q_{\theta}(x,dy)
    \right|
\end{align*}
bounded uniformly
because $q_\theta(x, dy)$ is the density of a Gaussian with mean $\meanq(x)/\theta$
and covariance $\covq(x)/\theta$,
and both $\meanq(x)$ and $\covq(x)$
are bounded uniformly in $x$ under Assumptions~\ref{def:model_setup}.
Furthermore, Condition~\ref{gamma_B_condition} of Assumptions~\ref{def:model_setup}
stipulates that $|\gamma(x, \eta) B^\theta_1(x, \eta)|$ is uniformly bounded,
and so recalling that $F$ is bounded above, we can now conclude that
under Assumptions~\ref{def:model_setup}
$\gamma(x, \eta) B^\theta_f(x, \eta) + f(x) F(x, \eta) \le C_f $ for some $C_f$.
